### 2024-06-03 journal/response.txt

We found it easier to combine Theorem 22 and Corollary 23 and their
proofs (now stated together as Theorem 22).

> line 1304, right column. "...will also keep the running time of our
> operations polylogarithmic *on $p$ and $q$*"

Done.

> Reviewer 2
>
> Then, the paper presents a modification of the first algorithm that
> uses bounded memory, more specifically polynomial on the number of
> processes and the size of the queue in the linearization, and with
> polylogarithmic amortized step complexity. As one would naturally
> expect, this version adds a sort of garbage collector mechanism that
> removes blocks of memory that are not needed anymore. I found this
> part of the paper harder to understand. I did not understand the need
> of using red-black trees, which I think complicates everything. I
> have no reason to think the implementation is wrong, all looks
> convincing to me, but I wonder if there is a way to simplify the
> presentation.

@@

> - ln 235, left. Repeated ‘words’.

Fixed.

> - For completeness, I suggest to add a preliminaries section, with a reminder of the main definitions such as linearizability, step complexity, and so on.

A new section on the model of computation has been added to include
these definitions.

@@

Done.

> - ln 845-848, right. I got confused here. Lemma 4 is only about the indexes, isn't it?

Since expression (3.3) defines subblocks in terms of indexes, we can apply Lemma 4.  Some more explanation has been added to the proof.

> - ln 1084-1088, right. Here do you mean that L is a valid sequential execution of the queue.

Technically, L is just a sequence of operations (without the
corresponding responses), rather than a sequential execution.
The wording has been clarified.

> - Lemma 15. After reading the proof I understood that legal linearization ordering only  means respecting real-time order. This is a concrete example why I think a preliminaries section might help.

The changes to the paragraph before Lemma 15 should help to clarify this.

> - ln 1192-1993, right. Please formalize the notion of time in an execution,

When revising, we were more careful about the use of "time".  (For
example, we now avoid using it when we mean step complexity.)  Since
the model section now defines an execution as a sequence of steps
(where the steps of processes are interleaved), some references to
time are now replaced by "step".  We have left some references to time
in informal descriptions, hoping that they will reinforce the intution
behind the arguments.

> - Theorem 22. What exactly is the size of the queue?

We added a clarification of what we mean just before the statement
of this theorem.

> - Sec 6. Why do you need read-back trees? How is that they facilitate the task of keeping memory bounded? Why the first algorithm cannot be modified to remove bocks, let us say using some sort of epoch-based memory reclamation scheme?

### 2024-06-03 journal/algorithm.tex
working on reviewer 2's comments
finished dealing with reviewer 2's comments

If a block contained an explicit representation of its sequences of \enqueues\ and \dequeues,
it would take $\Omega(p)$ time to construct a block, which would be too slow for task \ref{construct}.
Instead, the block stores an implicit representation of the sequences.
We now explain how we designed the fields for this implicit representation. 
Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{orderingtree} is actually represented, and Figure \ref{object-fields} for the definitions of the fields of \blocks\ and \nodes.

@@

A block is called the \emph{superblock} of all of its direct subblocks.
To facilitate navigating upwards in the ordering tree for task \ref{findinroot}, each block $B$ has a field \fld{super} that contains
the (approximate) index of its superblock in the parent \node's \fld{blocks} array (it may differ from the true index by 1).
%We shall ensure that the value of $B.\fld{super}$ differs from the true index of $B$'s superblock by at most 1.
This allows a process to determine the true location of the superblock by checking the \eleft\ or \eright\ values of just two \blocks\ in the parent \node.
Thus, starting from an operation in a leaf's block, one can use these indices to track the 
operation  up the path to the root, and determine the operation's location in a root block
in $O(\log p)$ time.

@@

Having defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
we can easily see how to construct a new block $B$ during a \op{Refresh} in $O(1)$ time.
A \op{Refresh} on \node\ $\var{v}$ reads the values $h_{\ell}$ and~$h_{r}$ of the \fld{head} fields of $\var{v}$'s children and stores 
$h_{\ell}-1$ and $h_{r}-1$ in $B.\eleft$ and $B.\eright$.
Then, we can compute 
\begin{eqnarray*}
B.\fld{sum\sub{enq}}&=&\var{v}.\fld{left}.\fld{blocks}[B.\eleft].\fld{sum\sub{enq}} \\
&&+ \var{v}.\fld{right}.\fld{blocks}[B.\eright].\fld{sum\sub{enq}}.
\end{eqnarray*}
For a block $B$ in the root, $B.\fld{size}$ is computed using the \fld{size} field of the previous block $B'$ and
the number of \enqueues\ and \dequeues\ in~$B$
\rev{as follows.  Before the operations of $B$, the queue has $B'.\fld{size}$ elements.
The \enqueues\ of $B$ add $B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}$ elements.
Then the \dequeues\ of $B$ remove $B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}$ elements.
(If subtracting this many elements would make the new size negative because there are more \dequeues\ than there are elements, then the new size is 0.)  Thus, we have}
\begin{eqnarray*}
B.\fld{size} &=& \max(0, B'.\fld{size} + (B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}})\\ 
			&& \hspace*{19mm}\mbox{ } - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}})).
\end{eqnarray*}

@@

\Function{Object}{Dequeue()}{} 
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,
	    \fld{sum\sub{enq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}$,\\
	    \fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1$}\label{deqNew}
    \State \Call{Append}{\var{B}}\Comment{\rev{append \dequeue\ to \var{leaf}.\fld{blocks} and propagate to root}}\label{deqAppend}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$} \Comment{\rev{find where \op{Deqeueue} is in \var{root}.\fld{blocks}}}\label{invokeIndexDequeue}
    \State \Return{\Call{FindResponse}{\var{b, i}}}\Comment{\rev{find response \dequeue\ should return}}\label{deqRest}
\EndFunction{Dequeue}

>>>>

If a block contained an explicit \rev{listing} of its sequences of \enqueues\ and \dequeues,
it would take $\Omega(p)$ \rev{steps} to construct a block, which would be too slow for task \ref{construct}.
Instead, the block stores an implicit representation of the sequences.
We now explain how we designed the fields for this implicit representation. 
Refer to Figure \ref{implicit} for an example showing how the tree in Figure \ref{orderingtree} is actually represented, and Figure \ref{object-fields} for the definitions of the fields of \blocks\ and \nodes.

@@

A block is called the \emph{superblock} of all of its direct subblocks.
To facilitate navigating upwards in the ordering tree for task \ref{findinroot}, each block $B$ has a field \fld{super} that contains
the (approximate) index of its superblock in the parent \node's \fld{blocks} array (it may differ from the true index by 1).
%We shall ensure that the value of $B.\fld{super}$ differs from the true index of $B$'s superblock by at most 1.
This allows a process to determine the true location of the superblock by checking the \eleft\ or \eright\ values of just two \blocks\ in the parent \node.
Thus, starting from an operation in a leaf's block, one can use these indices to track the 
operation  up the path to the root, and determine the operation's location in a root block
in $O(\log p)$ \rev{steps}.

@@

Having defined the fields required for tasks \ref{findinroot}, \ref{findrank} and \ref{findenqueue},
we can easily see how to construct a new block $B$ during a \op{Refresh} in $O(1)$ \rev{steps}.
A \op{Refresh} on \node\ $\var{v}$ reads the values $h_{\ell}$ and~$h_{r}$ of the \fld{head} fields of $\var{v}$'s children and stores 
$h_{\ell}-1$ and $h_{r}-1$ in $B.\eleft$ and $B.\eright$.
Then, we can compute 
\begin{eqnarray*}
B.\fld{sum\sub{enq}}&=&\var{v}.\fld{left}.\fld{blocks}[B.\eleft].\fld{sum\sub{enq}} \\
&&+ \var{v}.\fld{right}.\fld{blocks}[B.\eright].\fld{sum\sub{enq}}.
\end{eqnarray*}
For a block $B$ in the root, $B.\fld{size}$ is computed using the \fld{size} field of the previous block $B'$ and
the number of \enqueues\ and \dequeues\ in~$B$
\rev{as follows.  Before the operations of $B$, the queue has $B'.\fld{size}$ elements.
The \enqueues\ of $B$ add $B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}}$ elements.
Then the \dequeues\ of $B$ remove $B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}}$ elements.
(If subtracting this many elements would make the new size negative because there are more \dequeues\ than there are elements, then the new size is 0.)  Thus, we have}
\begin{eqnarray*}
B.\fld{size} &=& \max(0, B'.\fld{size} + (B.\fld{sum\sub{enq}}-B'.\fld{sum\sub{enq}})\\ 
			&& \hspace*{19mm}\mbox{ } - (B.\fld{sum\sub{deq}} - B'.\fld{sum\sub{deq}})).
\end{eqnarray*}

@@

\Function{Object}{Dequeue}{} 
    \State \hangbox{let \var{B} be a new \typ{Block} with fields \fld{element} \assign\ \nl,
	    \fld{sum\sub{enq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{enq}}$,\\
	    \fld{sum\sub{deq}} \assign\ $\var{leaf.}\fld{blocks}[\var{leaf.}\head-1].\fld{sum\sub{deq}}+1$}\label{deqNew}
    \State \Call{Append}{\var{B}}\Comment{\rev{append \dequeue\ to \var{leaf}.\fld{blocks} and propagate to root}}\label{deqAppend}
    \State $\langle \var{b}, \var{i}\rangle$ \assign\ \Call{IndexDequeue}{\var{leaf}, $\var{leaf.}\head-1$, $1$} \Comment{\rev{find where \op{Deqeueue} is in \var{root}.\fld{blocks}}}\label{invokeIndexDequeue}
    \State \Return{\Call{FindResponse}{\var{b, i}}}\Comment{\rev{find response \dequeue\ should return}}\label{deqRest}
\EndFunction{Dequeue}

### 2024-06-03 journal/analysis.tex
finished dealing with reviewer 2's comments

The following lemma helps bound the time for \op{GetEnqueue}.
\begin{lemma}\label{blockSize}
Each block $B$ in each node contains at most one operation of each process.
If $c$ is the execution's maximum point contention, $B$ has at most $c$ direct~subblocks.
\end{lemma}
\begin{proof}
Suppose $B$ contains an operation of process $p$.
Let $op$ be the earliest operation by $p$ contained in $B$.
When $op$ terminates, $op$ is contained in $B$ by Lemma \ref{lem::appendExactlyOnce}.
$B$ cannot contain any later operations by $p$, since $B$ is created before
those operations are invoked.

@@

We now bound step complexity in terms of the number of processes $p$, the maximum contention $c\leq p$, and the size of the queue.

>>>>

The following lemma helps bound the \rev{number of steps \op{GetEnqueue} takes}.
\begin{lemma}\label{blockSize}
Each block $B$ in each node contains at most one operation of each process.
If $c$ is the execution's maximum point contention, $B$ has at most $c$ direct~subblocks.
\end{lemma}
\begin{proof}
Suppose $B$ contains an operation of process $p$.
Let $op$ be the earliest operation by $p$ contained in $B$.
When $op$ terminates, $op$ is contained in $B$ by Lemma \ref{lem::appendExactlyOnce}.
$B$ cannot contain any later operations by $p$, since $B$ is created before
those operations are invoked.

@@

We now bound step complexity in terms of the number of processes $p$, the maximum contention $c\leq p$, and the size of the queue.
\rev{The bounds refer to the size of the queue when certain operations are linearized,
meaning the number of elements that would be in a queue after the operations in the prefix
of the linearization order up to that operation were performed sequentially.}.

### 2024-06-03 journal/correctness-GC.tex
finished dealing with reviewer 2's comments

\begin{customlemma}{\ref{lem::doubleRefresh}$'$}\label{lem::doubleRefreshGC}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}(\var{v}) by the same process.
All operations contained \var{v}'s children when $R_1$ begins
are contained in \var{v} when $R_2$ terminates.
\end{customlemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{casGC} or terminates at line \ref{addOPGC}, the claim follows
from Lemma \ref{successfulRefreshGC}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
Then some other \op{CAS} on $\var{v}.\fld{blocks}$ succeeds between the time each \op{Refresh}
reads $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC} and performs its \op{CAS} at line \ref{casGC}.
Consider the \op{Refresh} $R_3$ that does this successful \op{CAS} during $R_2$.
$R_3$ must have read $\var{v}.\fld{blocks}$ after the successful \op{CAS} during $R_1$.
The claim follows from Lemma \ref{successfulRefreshGC} applied to~$R_3$.
\end{proof}

@@

If \op{SplitBlock}($x$) returns the minimum block of $x$'s RBT at line \ref{substitute},
then the claim follows from the assumption that Invariant \ref{inv::finished} holds at all times before $st$.

@@

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains at most $3q_{max}+5p+1$ blocks.
\end{lemma}
\begin{proof}
Consider the invocation $A$ of \op{AddBlock} that updates a node \var{v}'s RBT with the insertion of a block whose 
\fld{index} is a multiple of $G$.
Then, $A$ performs a GC phase.
Let \rev{$t$ be the time} when $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
array.
Let $L_1$ be the prefix of the linearization $L$ corresponding to blocks $1..m$ of the root.
Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
$\ell$ is the last block added to the root's RBT before $t$.

@@

The number of \enqueues\ in $L_2$ whose values are still in the queue at the end of $L_2$ is at most $q_{max}$.
If the value enqueued by any \enqueue\ in $L_2$  is not still in the queue
at the end of $L_2$, then the \dequeue\ in $L_2$ that dequeued that value must still be in progress
at time $t$; otherwise the process that performed that \dequeue\ would have set its \var{last} entry
to the index of the root block that contains the \enqueue, which is greater than $m$, before $t$, contradicting
the fact that all values in the $last$ array at $t$ are less than or equal to $m$.
So, there are at most $p$ \enqueues\ in $L_2$ whose values are still in the queue at the end of $L_2$.
Thus, there are at most $q_{max}+p$ \enqueues\ in $L_2$.

If a \dequeue\ in $L_2$ returns a non-null value in the sequential execution $L$,
then the value it returns was either in the queue at the end of $L_1$ or it was enqueued during $L_2$.
Thus, there are at most $q_{max}+(q_{max}+p)$ \dequeues\ in $L_2$ that return non-null values.
Any \dequeue\ in $L_2$ that returns a null value in the sequential execution $L$ must still be in progress
at $t$; otherwise the process that performed the \dequeue\ would have set its \var{last} entry
to a value greater than the index of the root block that contains the \dequeue\ prior to $t$, 
contradicting the definition of $m$.
So, there are at most $p$ null \dequeues\ in $L_2$.
Thus, there are at most $2q_{max} + 2p$ \dequeues\ in $L_2$.

@@

We argue that each process $P$ can do an $r$th GC phase on \var{v} at most once.
Consider $P$'s first call $A$ to \op{AddBlock} that does an $r$th GC phase on \var{v}.
Let $\var{v},T,B$ be the arguments of $A$.
Any call to \op{AddBlock} on internal node \var{v} is from line \ref{refreshABGC} of \op{Refresh}, so $B.\fld{index}$
is the maximum \fld{index} in $T$ plus 1.
The \op{Refresh} that called $A$ performed a \op{CAS} at line \ref{casGC}.  Either the \op{CAS} succeeds
or it fails because 
some other \op{CAS} changes \var{v}.\fld{blocks} from $T$ to another tree.
Either way, by Lemma \ref{RBTupdates}, \var{v}'s RBT's maximum \fld{index} will be at least $r\cdot G$ at all times
after this CAS.
So if a subsequent \op{Refresh} by process $P$ 
ever calls \op{AddBlock} on \var{v} again, the block it passes as the third argument
will have $\fld{index}>r\cdot G$, so $P$ will not perform an $r$th GC phase on \var{v} again.

>>>>

\begin{customlemma}{\ref{lem::doubleRefresh}$'$}\label{lem::doubleRefreshGC}
Consider two consecutive terminating calls $R_1$, $R_2$ to \op{Refresh}(\var{v}) by the same process.
All operations contained \var{v}'s children when $R_1$ begins
are contained in \var{v} when $R_2$ terminates.
\end{customlemma}
\begin{proof}
If either $R_1$ or $R_2$ performs a successful \op{CAS} at line \ref{casGC} or terminates at line \ref{addOPGC}, the claim follows
from Lemma \ref{successfulRefreshGC}.
So suppose both $R_1$ and $R_2$ perform a failed \op{CAS} at line \ref{casGC}.
Then some other \op{CAS} on $\var{v}.\fld{blocks}$ succeeds between \rev{each \op{Refresh}'s 
read of $\var{v}.\fld{blocks}$ at line \ref{refreshReadTGC} and its \op{CAS}} at line \ref{casGC}.
Consider the \op{Refresh} $R_3$ that does this successful \op{CAS} during $R_2$.
$R_3$ must have read $\var{v}.\fld{blocks}$ after the successful \op{CAS} during $R_1$.
The claim follows from Lemma \ref{successfulRefreshGC} applied to~$R_3$.
\end{proof}

@@

If \op{SplitBlock}($x$) returns the minimum block of $x$'s RBT at line \ref{substitute},
then the claim follows from the assumption that Invariant \ref{inv::finished} holds before $st$.

@@

\begin{lemma}\label{boundAfterGC}
If the maximum \fld{index} in a node's RBT is a multiple of $G$, then
it contains at most $3q_{max}+5p+1$ blocks.
\end{lemma}
\begin{proof}
Consider the invocation $A$ of \op{AddBlock} that updates a node \var{v}'s RBT with the insertion of a block whose 
\fld{index} is a multiple of $G$.
Then, $A$ performs a GC phase.
Let \rev{$s$ be the step} when $A$ invokes \op{SplitBlock} on line \ref{callSplitBlock}.
That call to \op{SplitBlock} recurses up to the root, where it computes $m$ by reading the \var{last}
array.
Let $L_1$ be the prefix of the linearization $L$ corresponding to blocks $1..m$ of the root.
Let $L_2$ be the next segment of the linearization corresponding to blocks $m+1..\ell$ of the root, where
$\ell$ is the last block added to the root's RBT before $s$.

@@

The number of \enqueues\ in $L_2$ whose values are still in the queue at the end of $L_2$ is at most $q_{max}$.
If the value enqueued by any \enqueue\ in $L_2$  is not still in the queue
at the end of $L_2$, then the \dequeue\ in $L_2$ that dequeued that value must still be in progress
when $s$ occurs; otherwise the process that performed that \dequeue\ would have set its \var{last} entry
to the index of the root block that contains the \enqueue, which is greater than $m$, before $s$, contradicting
the fact that all values in the $last$ array are less than or equal to $m$ when $s$ occurs.
So, there are at most $p$ \enqueues\ in $L_2$ whose values are still in the queue at the end of~$L_2$.
Thus, there are at most $q_{max}+p$ \enqueues\ in~$L_2$.

If a \dequeue\ in $L_2$ returns a non-null value in the sequential execution $L$,
then the value it returns was either in the queue at the end of $L_1$ or it was enqueued during $L_2$.
Thus, there are at most $q_{max}+(q_{max}+p)$ \dequeues\ in $L_2$ that return non-null values.
Any \dequeue\ in $L_2$ that returns a null value in the sequential execution $L$ must still be in progress
when $s$ occurs; otherwise the process that performed the \dequeue\ would have set its \var{last} entry
to a value greater than the index of the root block that contains the \dequeue\ prior to $s$, 
contradicting the definition of $m$.
So, there are at most $p$ null \dequeues\ in $L_2$.
Thus, there are at most $2q_{max} + 2p$ \dequeues\ in $L_2$.

@@

We argue that each process $P$ can do an $r$th GC phase on \var{v} at most once.
Consider $P$'s first call $A$ to \op{AddBlock} that does an $r$th GC phase on \var{v}.
Let $\var{v},T,B$ be the arguments of $A$.
Any call to \op{AddBlock} on internal node \var{v} is from line \ref{refreshABGC} of \op{Refresh}, so $B.\fld{index}$
is the maximum \fld{index} in $T$ plus 1.
The \op{Refresh} that called $A$ performed a \op{CAS} at line \ref{casGC}.  Either the \op{CAS} succeeds
or it fails because 
some other \op{CAS} changes \var{v}.\fld{blocks} from $T$ to another tree.
Either way, by Lemma \ref{RBTupdates}, \var{v}'s RBT's maximum \fld{index} will always be at least $r\cdot G$ 
after this CAS.
So if a subsequent \op{Refresh} by process $P$ 
ever calls \op{AddBlock} on \var{v} again, the block it passes as the third argument
will have $\fld{index}>r\cdot G$, so $P$ will not perform an $r$th GC phase on \var{v} again.
